Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-04 paper/paper.tex
methods

\subsection{Setting up the Inference}
 \emph{Think about presentation, prioritizing/highlighting prior but logically building it from an full inference perspective and needing a prior, then dive into prior stuff. It's confusing right now jumping between inference and prior building !!}

@@

Equation~\ref{eq:bayes} is too simplistic to describe this procedure.
Because our prior is the \cmd, our inference will include not only the parallax from \gaia, but also the photometry of the stars. Specifically, we use the $J$ and $K$ band photometry from \tmass, as well as the parallax from \gaia. Our true, latent values are the $(J-K)^{true}_n$ color and the $J$ band absolute magnitude $M^{true}_{J,n}$ for each star, which we encapsulate in the true vector $\mathbf{Y}$. The observed data is the dust corrected color $(J-K)^c_n$, the \gaia\ parallax $\varpi_n$ in units of $[mas]$, and the dust corrected $J$ band magnitude $J^c_n$, which we encapsulate in the data vector $\mathbf{y}$.

We now describe a singularity of our method, which allows us to conservatively deconvolve the data into a \cmd\ prior without requiring a full hierarchical inference framework (which was explored in \citealt{leistedtHogg2017}).
Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\mathbf{y}$ contains a transformation of the dust corrected absolute magnitude
\begin{equation}
y_0 = \varpi\,10^{0.2J^c_n} = 10^{0.2\,M^c_{J,n} + 2}
\label{eq:transform}
\end{equation}
The transformation of the dust corrected absolute magnitude is required to to keep the uncertainties Gaussian, which, under some assumptions along with this transformation, we maintain. Gaussian uncertainties are required to generate our prior using \xd, but more about this and our dust corrections below.
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the 3D dust map. So our full terminology is
\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\mathbf{Y}_n = [10^{0.2\,M^{true}_{J,n} + 2}, (J-K)^{true}_n] \\
\mathrm{data \, vector} \quad &\mathbf{y}_n = [\varpi_n\,10^{0.2J^c_n}, (J- K)^c_n] \;\hspace*{15mm}  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^c_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J - K)^c_n = J_n - K_n - Q_{JK}\,E(B-V)_n \\
\mathrm{under\ the\ assumptions} \quad &I_n = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}
where $E(B-V)_n$ is the output of the dust model of \cite{green15}, and $Q_{\lambda}$ is the correction factor for a photometric band assuming some reddening law.

We assume that the parallax likelihood is Gaussian, as well as the $(J-K)^c_n$ color likelihood. So our full likelihood of the $n$th star is
\begin{equation}
p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) = \mathcal{N}(\mathbf{y}_n \given \mathbf{Y}_n, V_n)
\end{equation}
with \\
\[
V_n = \begin{bmatrix}
\sigma_{J,n}^2 + \sigma_{K,n}^2 & 0 \\
0 & (\sigma_{\varpi}\,10^{0.2\,J_n^C})^2
\end{bmatrix}
\]
where $\mathcal{N}$ is a 2D Gaussian function with mean $\mathbf{Y}$ and variance $V_n$.
Here we have assumed that the parallax uncertainty is significantly larger than the $J$ band uncertainty, and therefore the dominant uncertainty in the transformed absolute magnitude. So we take $J_n^C$ as a point estimate and only propagate the parallax uncertainty.

We can now write the multidimensional analogous to Equation \ref{eq:bayes},
\begin{equation}
p(\mathbf{Y}_n \given \mathbf{y}_n, I_n) = p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) \, p(\mathbf{Y}_n) \label{eq:posterior}
\end{equation}
where $p(\mathbf{y}_n \given \mathbf{Y}_n, I_n)$ is the likelihood of the data vector, and $p(\mathbf{Y}_n)$ is our prior.
Before we describe how this can be turned into a posterior belief on the true parallax, we turn our attention to the prior, which will be generated from the data themselves. We will use XD, which will take advantage of the Gaussian likelihood.

@@

To generate our prior $p(\mathbf{Y}_n)$, we fit our data in the slightly transformed \cmd\ space using \xd, which deconvolves the data $\{ \mathbf{Y}_n \}$ with its uncertainties $\{ V_n\}$, which are Gaussian, heterogeneous measurement uncertainties. It generates an estimate of the underlying distribution from which the uncertain data were drawn from, modeled as a mixture of $K$ Gaussians:
\begin{equation}
	p\bigl(\ \mathbf{Y} \given  \left\{\mathrm{amp}_k, \mathbf{\mu}_k, V_k\right\}_{k=1}^K \bigr) = \sum_{k=1}^K \mathrm{amp}_k \ \mathcal{N}(\mathbf{Y} \given \mathbf{\mu}_k, V_k )
\end{equation}
where $\mathbf{\mu}_k$ and $V_k$ are a two dimensional vector and matrix, respectively, which correspond to the location and covariance of the $k$th component (a Gaussian in our transformed \cmd\ space). The relative amplitudes between the components are described by the scalars $\{ \mathrm{amp}_k \}$.

Given the data $\{ \mathbf{Y}_n, V_n\}$, \xd\ maximizes the marginalized likelihood of the data, marginalizing over the true data. In equations, it finds the coefficients
\begin{eqnarray}
\alpha &=& \left\{\mathrm{amp}_k, \mathbf{\mu}_k, V_k\right\}_{k=1}^K \nonumber\\
       &=& \argmax_{\alpha} \, \Pi_n \, p(\mathbf{y}_n \given \alpha) \nonumber\\
       &=& \argmax_{\alpha} \, \Pi_n \, \int p(\mathbf{y}_n \given \mathbf{Y}_n) \, p(\mathbf{Y}_n \given \alpha)\mathrm{d\mathbf{Y}_n}
\label{eq:xdmml}
\end{eqnarray}
which is made analytically tractable with the Gaussian likelihood and mixture-of-Gaussians prior \citep{bovy11}.

>>>>

\subsection{Setting up the Inference}

@@

Equation~\ref{eq:bayes} does not fully specify this procedure.
Because our prior comes from the \cmd, our inference will include not only the parallax from \gaia, but also the photometry of the stars. Specifically, we use the $J$ and $K$ band photometry from \tmass, as well as the parallax from \gaia. Our true, latent values are the $(J-K)^{true}_n$ color and the $J$ band absolute magnitude $M^{true}_{J,n}$ for each star, which we encapsulate in the true vector $\mathbf{Y}$. The observed data is the dust corrected color $(J-K)^c_n$, the \gaia\ parallax $\varpi_n$ in units of $[mas]$, and the dust corrected $J$ band magnitude $J^c_n$, which we encapsulate in the data vector $\mathbf{y}$.

We now describe a singularity of our method, which allows us to conservatively deconvolve the data into a \cmd\ prior without requiring a full hierarchical inference framework (which was explored in \citealt{leistedtHogg2017}).
Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\mathbf{y}$ contains a transformation of the dust corrected absolute magnitude
\begin{equation}
y_0 = \varpi\,10^{\frac{1}{5}\,J^c_n} = 10^{\frac{1}{5}\,M^c_{J,n} + 2}
\label{eq:transform}
\end{equation}
where, for this expression, it is important to remember that $\varpi$ is assumed to be in units of $[mas]$.
The transformation of the dust corrected absolute magnitude is required to to keep the uncertainties Gaussian, which, under some assumptions along with this transformation, we maintain (Gaussian uncertainties are required to use \xd, see below).
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the 3D dust map. So our full terminology is
\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\mathbf{Y}_n = [10^{0.2\,M^{true}_{J,n} + 2}, (J-K)^{true}_n] \\
\mathrm{data \, vector} \quad &\mathbf{y}_n = [\varpi_n\,10^{0.2J^c_n}, (J- K)^c_n] \;\hspace*{15mm}  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^c_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J - K)^c_n = J_n - K_n - Q_{JK}\,E(B-V)_n \\
\mathrm{under\ the\ assumptions} \quad &I_n = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}
where $E(B-V)_n$ is the output of the dust model of \cite{green15}, and $Q_{\lambda}$ is the correction factor for a photometric band assuming some reddening law.

We assume that the parallax likelihood is Gaussian, as well as the $(J-K)^c_n$ color likelihood. So our full likelihood for the $n$th star is
\begin{equation}
p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) = \mathcal{N}(\mathbf{y}_n \given \mathbf{Y}_n, V_n)
\end{equation}
with \\
\[
V_n = \begin{bmatrix}
\sigma_{J,n}^2 + \sigma_{K,n}^2 & 0 \\
0 & \left(\sigma_{\varpi}\,10^{0.2\,J_n^C}\right)^2
\end{bmatrix}
\]
where $\mathcal{N}(\boldsymbol{\mu}, \mathbf{\Sigma})$ represents a normal distribution with mean vector $\boldsymbol{\mu}$ and covariance matrix $\mathbf{\Sigma}$.
Here we have assumed that the parallax uncertainty is significantly larger than the $J$-band magnitude uncertainty and is therefore the dominant uncertainty in the transformed absolute magnitude. We take $J_n^C$ as a point estimate and only propagate the parallax uncertainty.

We can now write the multidimensional posterior expression analogous to Equation~\ref{eq:bayes},
\begin{equation}
p(\mathbf{Y}_n \given \mathbf{y}_n, I_n) \propto p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) \, p(\mathbf{Y}_n) \label{eq:posterior}
\end{equation}
where $p(\mathbf{y}_n \given \mathbf{Y}_n, I_n)$ is the likelihood of the data vector, and $p(\mathbf{Y}_n)$ is our prior.
Before we describe how this can be turned into a posterior belief on the true parallax, we turn our attention to the prior, which will be generated from the data themselves. For this, we use XD, which takes advantage of the Gaussian likelihood.

@@

To generate the empirical prior $p(\mathbf{Y}_n)$, we fit our data in the transformed \cmd\ space (see previous sub-section) using \xd, which deconvolves the data $\{ \mathbf{Y}_n \}$ with heterogeneous, Gaussian noise variances $\{ V_n\}$. \xd\ generates an estimate of the underlying distribution from which the uncertain data were drawn, modeled as a mixture of $K$ Gaussians:
\begin{equation}
	p\bigl(\ \mathbf{Y} \given  \left\{A_k, \mathbf{\mu}_k, V_k\right\}_{k=1}^K \bigr) = \sum_{k=1}^K A_k \ \mathcal{N}(\mathbf{Y} \given \mathbf{\mu}_k, V_k )
\end{equation}
where $\mathbf{\mu}_k$ and $V_k$ are a two dimensional vector and matrix, respectively, which correspond to the location and covariance of the $k$th component (a Gaussian in our transformed \cmd\ space). The relative amplitudes between the components are described by the scalars $\{ A_k \}$.

Given the data $\{ \mathbf{Y}_n, V_n\}$, \xd\ maximizes the marginalized likelihood of the data, marginalizing over the true data. Assuming $\alpha = \left\{A_k, \mathbf{\mu}_k, V_k\right\}_{k=1}^K$, it finds the coefficients
\begin{eqnarray}
\hat{\alpha} &=& \argmax_{\alpha} \, \prod_n \, p(\mathbf{y}_n \given \alpha) \nonumber\\
             &=& \argmax_{\alpha} \, \prod_n \, \int p(\mathbf{y}_n \given \mathbf{Y}_n) \, p(\mathbf{Y}_n \given \alpha)\mathrm{d\mathbf{Y}_n}
\label{eq:xdmml}
\end{eqnarray}
which is made analytically tractable with the Gaussian likelihood and mixture-of-Gaussians prior \citep{bovy11}.
